---
title:Tidyverse I: tidy data
webr:
  resources:
    - ../data/dutch_demographics.csv
    - ../data/dutch_elections_2023.csv
    - ../data/shapes_nl.rds
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ../_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| autorun: true
#| include: false
library(tidyverse)
library(dslabs)
gap <- as_tibble(gapminder)
gap_asia <- filter(gap, continent == "Asia")
head(gap_asia)
```


## Introduction to Tidyverse

The goal of this chapter is to get you acquainted with the
[Tidyverse](https://www.tidyverse.org/). Tidyverse is a collection of
packages that have been designed around a singular and clearly defined
set of principles about what data should look like and how we should
work with it. It comes with a nice introduction in the [R for Data
Science](https://r4ds.hadley.nz/) book, for which the digital version is
available for free. This chapter deals with most of the material in
chapter 5 of that book.

In this chapter, we’ll focus on working with data using the
`tidyverse` package. This package includes the `dplyr` (data-pliers)
packages, which contains most of the tools we’re using below, but it
also contains functions for reading, analysing and visualising data that
will be explained later.

### The data challenge: Mind the gap!

For this example, let's have a look at the `gapminder` dataset,
which contains collection of worldwide, long-term statistics on social, economic, and environmental development.

This dataset is available in R directly in the `dslabs` (data science labs) package. Note that if you run this on your own computer you may have to install the package first with `install.packages("dslabs")`.
Let's convert the data to a tidyverse 'tibble' (just so it displays a bit nicer) and look at the top rows:

```{webr}
#| warning: false
library(dslabs)
library(tidyverse)
gap <- as_tibble(gapminder)
head(gap)
```


::: {.challenge}
Suppose we would like to know the relation between GDP per capita and fertility in Asian countries.
Can you think of the sequence of steps that we would need to get from this dataset
to an answer to that question?

Don't worry if you can't see the answer yet -- this chapter will take you through the steps
to prepare the data and answer that question. Once you have all the tools in your belt,
you will be able to answer questions like this.
:::

::: {.callout-note collapse="true" title="Columns in the gapminder data - click to expand"}

| Variable         | Explanation                                          |
|------------------|------------------------------------------------------|
| country          | Name of the country                                  |
| year             | Year of observation                                  |
| infant_mortality | Infant deaths per 1,000 live births                  |
| life_expectancy  | Life expectancy at birth, in years                   |
| fertility        | Average number of children per woman                 |
| population       | Total population                                     |
| gdp              | Gross domestic product                               |
| continent        | Continent the country belongs to                     |
| region           | Geographic region within the continent               |

:::


## The tidyverse package

Taking a step back, tidyverse is a package (or actually a collection of packages).
This means that it's not included in R, but you need to install it separately.  You've probably already done that, but if not,
install it with the *packages* pane in Rstudio or call `install.packages("tidyverse")`. This is only needed once on your computer, so you don't need to include this code in your scripts.

What you do need to do every time is *activate* the package using
the `library` function:

```{r, eval=F}
library(tidyverse)
```

::: {.callout-note title="Common errors"}
The two most common mistakes in R are forgetting to install a package and forgetting to activate it. In both cases, you will get an *error message*: a text (often displayed in red) that explains what went wrong. Don't panic if you see an error: you will quickly learn to recognize the most common errors, and even if you don't, copy pasting the error into a search or chat engine (or an email to your instructor) often gives you a quick solution.

As an example, try the following two calls:

```{r, eval=F}
library(nastyverse)
format_log(100)
```

Both will give you an error message! The first will tell you `there is no package called 'nastyverse'` -- this means that either you made a typing error, or you forgot to install the package. If you think the name is correct, a quick `install.packages` generally fixes the problem.

The second will say `could not find function "format_log"`. Again, unless you made a typo, this generally means you forgot to activate a package, in this case the `scales` package, which you can fix with `library(scales)`. But the most important thing is to not panic, read the error message, and either fix it or search for a solution.

**Asking for help:** Also, remember that error messages are actually meant to be helpful, even if they sound complicated and technical.
The best thing to do is ask for help, including the command you tried and the full error message.
You can copy paste this into ChatGPT (or Gemini, Claude, etc) or just into google, and it will often tell you  what the problem is and how to fix it. If you have one, you can of course also ask your instructor.
By including the command and full error message, they will be able to help you much better than if you
only say that "it doesn't work".
:::

### Data Frames


In R, most data is in what we call a `data frame`:
a rectangular data set consisting of rows
(often items or respondents) and columns (often measurements of or data
about these items). In 'vanilla' R, these are called `data.frame`, but
tidyverse has its own version of data frames called `tibble`, which will
be the default type we use in this book.

R and tidyverse have a number of ways to inspect the shape and contents of a data frame.
We've already seen `head` used above, to print the first few rows of a data frame.
In fact, when you just run the name of a tibble, it also prints the top rows (and first few columns).
This is one of the advantages of tibbles over default ('vanilla') data frames:
they detect the size of your screen and print in a way that fits the screen nicely.
In general, we advise using `head(data)` rather than just `data`,
because in a Quarto report it will otherwise print the enitre data set, which is normally not what you want.

Another really useful function is `glimpse`, which gives an overview of all the columns in the data and a sample of values:

```{webr}
glimpse(gap)
```

As you can see when you run this, it displays the total number of rows and columns, and for each column shows the name, type, and a sample of some values.
If you're working in RStudio, you can also use `View(data)`, which opens the data in a spreadsheet-like view.

::: {.callout-note title="Understanding Functions"}

The main way to transform and analyse data is by calling *functions*.
In fact, we've done this a number of times before, such as when we called
`as_tibble(gapminder)` or `head(gap)`.
You can recognize a function by the parentheses: `head(gap)` calls the function `head` on the data `gap`,
displaying the top rows (the 'head' of the data).

Tidyverse offers a number of functions that allow you to manipulate data frames.
In general, a function can be seen as a command or instruction to the computer to do something.
Tidyverse functions almost always take the data frame as a first argument, and *return* a transformed
copy of the data. To store this transformed copy, we generally *assign* it to a new or existing object,
such as when we assigned the result of `as_tibble(gapminder)` to the new object `gap`:

```{r, eval=F}
gap <- as_tibble(gapminder)
```

In fact, most of our code looks like this: call a function on a data set, and assign it to a new or existing name.
This chapter will introduce you to a number of tidyverse functions, for example for filtering and sorting data.
:::

## Managing columns with `select` and `rename`

Let's return to the `gapminder` data set.
Many datasets contain many more columns than you need,
and sometimes also have column names that are hard to understand (like `q12`)
or hard to use in R code because they contain spaces or start with numbers.
So, let's look at how to select and rename columns.

### Selecting columns with `select`

In data cleaning, a first step is often to *select* only the columns that you need,
making it easier to see what's going on in the data.
For example, if we only want to select the country, year and gdp,
we can select these columns as follows:

```{webr}
d <- select(gap, country, year, gdp)
head(d)
```

As you can see, we use the standard pattern here of `result <- verb(data, options)`:
we *select* the specified columns from *gap*, and assign the result to *d*.
Note that `d` is simply a name *you* assign to the result, so you can pick any name you want
(and often, a more descriptive name is a good idea).

`select` has a couple more tricks up its sleeve: you can select column ranges
by using `first_column:last_column`. For example, `select(gap, country:fertility)` will select the first five columns from the original data. You can also drop one or more columns by using a minus sign:
`select(gap, -region)` will keep all columns except for region.

### Renaming columns with `rename`

There are two ways to rename columns.
First, the `rename` command does just that: `rename(data, new_name=old_name)` will rename the `old_name` column into `new_name`. For example, `rename(gap, GDP=gdp)` will capitalize the GDP column.

The second method is very similar: in select, you can also use `new_name=old_name` to rename a column:
`select(gap, country, year, GDP=gdp)` will select the columns country, year, and GDP (renamed from gdp).
The difference is that select will drop all the other columns, while rename will only rename the mentioned columns but leave the rest as is.

::: {.callout-note title="Illegal names"}

Renaming is especially useful for column names that cannot be directly used in R.
Suppose that the `life_expectancy` column would instead be called `Life Expectancy` (with a space).
If you tried to call `select(gap, Life Expectancy)`, R would throw an error because it doesn't understand that `Life Expectancy` is supposed to be a single name.
You can work around this by placing backticks (`` ` ``) around the column name: ``select(gap, `Life Expectancy`)``.
However, that gets old really fast, especially because the backtick is very often hard to find on a keyboard.

So, the general advice is to *rename* columns that are not a legal name, e.g. by replacing spaces by underscores (`_`) or by just using a different name. In the rename (or select) command, you can place either backticks or normal quotes (`"` or `'`) around the old name. Suppose that `gap_dirty` contains the column with a space, you can simply use quotes in `select` or `rename` to rename it:

```{r, eval=F}
gap_cleaned <- rename(gap_dirty, life_expectancy = "Life Expectancy")
# or alternatively:
gap_cleaned <- select(gap_dirty, year, country, life_expectancy = "Life Expectancy")
```

:::

### Assigning results to objects {#sec-assigning}

In the examples above, the result of the select or rename was
assigned to a new object, `d` (and later `gap_cleaned`).
The original data `gap` remains the same, and `d` contains the result of the function.

This is a common pattern across tidyverse functions:
they don't really change the data, they return a *changed copy* of the data.
This gives you three options for handling the result.

If you assign it to a new object, both data frames are kept and you can later
work with either:


```{r, eval=F}
new_data <- select(gap, country, year, gdp)
```

You can also re-assign it to the same object.
This essentially replaces the object with the transformed version,
meaning only the new version is now available:

```{r, eval=F}
gap <- select(gap, country, year, gdp)
```

Finally, you can choose not to assign it to anything.
This will cause the result to be printed on the screen, but not remembered:

```{r, eval=F}
select(gap, country, year, gdp)
```

In general, if you only need the new version it's usually best to just replace
the object, since otherwise you clutter your workspace with multiple versions of the same data.
Assign to a new object if you need to do specific analyses on e.g. a subset of the data,
but later want to return to the original for other analyses.
Finally, if you only want to inspect the data there is no need to assign it to anything.

### Exercise: selecting and renaming columns

{{< include exercises/_ch03_ex_select.qmd >}}

## Subsetting data with `filter`

In the previous section we looked at selecting columns with the `select`
function. Another common task is selecting *rows*, which can be done
with the `filter` function.

### The `filter` function

Looking back at the original question, we would like to determine
the relationship between GDP and fertility in Asian countries.
A logical cleaning step would be to make a subset of the data
that only contain rows where `continent` equals `"Asia"`.
This can be achieved with the following code:


```{webr}
library(dslabs)
library(tidyverse)
gap <- as_tibble(gapminder)
gap_asia <- filter(gap, continent == "Asia")
head(gap_asia)
```

The line that does the filtering can be read as:
"Filter the 'gap' dataset to keep only the rows where the `continent` column equals
the text `"Asia"`, and assign the result to `gap_asia`".

::: {.callout-note title="Why the double equal sign (==)?"}

Note that we use a double equals sign (`==`): in R,
when you ask a **question** ("are two things equal?") you always use `==`.

The single equals sign (`=`) means **assignment** — giving something a name or value.
You've already seen this with named arguments to functions:
`rename(gap, GDP=gdp)` doesn't ask whether `gdp` equals `GDP`,
it assigns the column `gdp` to the new name `GDP`.

Note: You may also see `=` used to assign variables, e.g. `x = 5`. This works and is identical to `x <- 5`. However, `<-` is more idiomatic in R and will be used throughout this book.
:::

### Other Comparisons

Besides checking for equality, we can also filter on a value being larger or smaller than
(or just different from) another value. This is achieved using different *operators* to compare
the values. The most common comparison operators in R are:

| Operator | Meaning                  | Example                  | Reads as                                        |
|----------|--------------------------|--------------------------|-------------------------------------------------|
| `==`     | equal to                 | `continent == "Asia"`    | continent is exactly `"Asia"`                   |
| `!=`     | not equal to             | `continent != "Asia"`    | continent is anything other than `"Asia"`       |
| `<`      | less than                | `year < 1980`            | year is before 1980 (excluding 1980)            |
| `>`      | greater than             | `gdp > 1000`             | gdp is above 1000 (excluding 1000)              |
| `<=`     | less than or equal to    | `year <= 1980`           | year is 1980 or earlier                         |
| `>=`     | greater than or equal to | `life_expectancy >= 70`  | life expectancy is 70 or higher                 |
| `%in%`     | one of the values        | `continent %in% c("Asia", "Europe")` | Continent is one of the values `"Asia"` or `"Europe"` |

So, for example you can select only data from countries with a population of at least 10 million using the `>=` operator:


```{webr}
filter(gap_asia, population >= 10000000)
```

Note: In the code above the result of `filter` is not assigned to an object. This means that the result is shown on the screen, but not stored anywhere. This can be very useful to quickly inspect something, but remember that you need to store the result if you want to use it later. See @sec-assigning above for more detail.

::: {.callout-note title="To quote or not to quote"}
In the code above, we 'quoted' the term `Asia` by surrounding it with quotation marks (`"`).
At the same time, we didn't quote numbers like `10000000` or `gap` or `continent`. This may seem confusing, but there is method to this madness:

- You need to use quotes for *literal text*, such as the word `"Asia"`
- You do not use quotes for *numbers*, such as `10000000`
- You do not use quotes for *names* of data frames or columns, such as `gap` or `continent`

There is one exception to the last rule: if the column is not a legal value for a name, e.g. because it contains a space or starts with a number, you need to quote it using backticks (`` ` ``). However, like we explained above, in the select and rename commands you can also use normal quotes to refer to a column containing spaces.
:::

### Combining criteria

Finally, you can combine multiple criteria in a single filter call.
This can be achieved using `&` meaning *and*, and `|` meaning *or*.
If you combine both `&` and `|` in a single call, you should use parentheses
to make clear what you mean:
`(year == 2015 | year == 2016) & region == "Southern Asia"` will select all entries from Southern Asia where the year is either 2015 or 2016. If you place the parentheses differently, the meaning shifts: `year == 2015 | (year == 2016 & region == "Southern Asia")` means selecting all data from the year 2015, and data from 2016 only where the region is Southern Asia.

If you give multiple criteria as separate arguments (that is, separated with a comma `,`)
they effectively function as *and* as well. So, `filter(year==2015, country=="Brunei")` is identical to `filter(year == 2015 & country == "Brunei")`, and will only keep the rows for that specific country in that specific year.

Finally, you can also use `!` to negate a condition. For example, `!(continent %in% c("Asia", "Europe"))` keeps rows where the continent is not Asia or Europe. Together, `&`, `|`,  `!`, and `()` let you build up arbitrary combinations of conditions."


### Exercise: Filtering data

{{< include exercises/_ch03_ex_filter.qmd >}}

### Removing missing values

Often, data contains *missing values*, for example survey questions that
were not answered or other data that is missing or unknown.
In the `gapminder` data, for example, the data for infant mortality, fertility, population and GDP is sometimes missing:

```{webr}
#| autorun: true
gap_asia
```

In R, such
values are called `NA` (Not Available).
Many operations in R, such as calculating descriptive statistics or running models, can give unexpected results if the data contain missing values.
For this reason, you often have to deal with missing values explicitly, often by removing the rows that are missing crucial data.

A complication here is that comparisons with
missing values are automatically missing as well:
if `gdp` is missing, `gdp > 100` will also be `NA`.
This may seem strange, but the reason for this is that we don't know whether GDP is larger or smaller than 100, as we are missing that measurement. Even `gdp == NA` does not work: if gdp is missing, we don't know whether it's equal to some other value that's also missing, so the comparison will always result in `NA`.

For this reason, you need to use special functions to deal with missing values: `is.na(column)` tests which values in a specific column are missing,
while `drop_na(data)` removes data with missing values.

Going back to the example, to see which rows have missing gdp data we can use `filter(is.na(gdp))`:

```{webr}
filter(gap_asia, is.na(gdp))
```

To remove all rows where gdp is missing, you can use  `filter(!is.na(gdp))`, but you can also use the special purpose function `drop_na`:

```{webr}
drop_na(gap_asia, gdp)
```

Note that if you provide multiple columns to `drop_na` it drops rows where any of the columns is missing. If you don't give any columns it will drop all rows with missing values (the equivalent to the base-R `na.omit` function).


{{< include exercises/_ch03_ex_filter_na.qmd >}}

## Sorting data with `arrange`

The output in the examples above were all shown in the sort order the data originally had, which turns out to be by year (starting with 1960) and then by country. In many cases, you want to sort by value so you can immediately see e.g. which countries have the highest life expectancy.

### The arrange function

This can be achieved with the `arrange` function:

```{webr}
arrange(gap_asia, life_expectancy)
```

So, it turns out that Cambodia had extremely low life expectancy during the Khmer Rouge regime in the late 1970's (and if you look at the population column you can see total population actually dropped from over 7.5 million in 1975 to under 7 million from 1978!)

By default, `arrange` sorts in ascending order (low to high or A-Z).
To switch the sorting order, you can use the `desc(column)` function:

```{webr}
arrange(gap_asia, desc(life_expectancy))
```

This shows Hong Kong and Japan in the last decade of the data set to have the highest life expectancy of 83 years.

Sorting also works for text columns. So, to sort by country name starting from z, you could use:

```{webr}
arrange(gap_asia, desc(country))
```


### Multiple sort columns

Finally, you can include multiple sort columns by just giving them as additional arguments.
For example, suppose we want to list first by year (descending) and then (for rows within the same year) by life expectancy, we can use the following:

```{webr}
arrange(gap_asia, desc(year), life_expectancy)
```

So, in 2016 the lowest life expectancy was in Yemen at almost 65 years.

Note that missing values are always sorted at the bottom. So,
neither `arrange(gap_asia, gdp)` nor `arrange(gap_asia, desc(gdp))`
will show the missing values in the gdp column at the top.
To view missing values, you need to do a (temporary) filter to select the missing values using `filter(is.na(column))` as explained above.

### Exercise

{{< include exercises/_ch03_ex_arrange.qmd >}}


## Changing or computing values

As a final part of this tutorial, let’s look at how you can change
values using the `mutate` function:

### The `mutate()` function

The mutate function makes it easy to create new variables or to modify
existing ones. For those more familiar with SPSS, this is what you would
do with `compute` and `recode`. As a simple example, let’s start again
with the 538 gun polls data used above:

``` r
library(tidyverse)
url <- "https://raw.githubusercontent.com/fivethirtyeight/data/master/poll-quiz-guns/guns-polls.csv"
gunpolls <- read_csv(url)
gunpolls <- select(gunpolls, Question, Support,
                   rep=`Republican Support`, dem=`Democratic Support`)
```

For example, to transform the Support percentage (0-100) to a proportion
(0-1), we can divide the column by 100 using `mutate`:

``` r
mutate(gunpolls, Support_Prop = Support / 100)
```

The syntax of mutate is similar to that of `filter()` and `select()`:
The first argument is the data frame to mutate, and then any number of
additional arguments can be given to perform mutations. The mutations
themselves are named arguments, in which you can provide any
calculations using existing columns such as `Support`. The result of the
calculatoin will then be stored in the name of the argument, in this
case `Support_Prop`. You can also overwrite columns by storing the
result with an existing column’s name.

For calculations, you can use mathematical symbols such as `+`, `/`
(division), and `*` (multiplication). There are also many useful
functions that you can use in calculations, such as `round()` to round
values, `as.numeric()` to change a text column to a numberic column, or
`abs()` to take the absolute (positive) value of numbers. Feel free to
play around a bit in the code above.

### Exercise

Using the `gunpolls` data as defined above, can you create a new column
`party_diff` that contains the absolute value of the difference between
republican and democratic support? That is, if republican support is 61
and democratic support is 86 (or the other way around), the difference
should be 25. Overwrite the gunpolls object with the result (i.e. add
the column to the existing data set).

``` r
gunpolls <- mutate(gunpolls, ____)
gunpolls
```
