---
title: "Reshaping data (todo)"
webr:
  resources:
    - ../data/temperature_anomaly.csv
    - ../data/co2_emissions.csv
    - ../data/income_groups.csv
editor_options: 
  chunk_output_type: console
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}
{{< include ../_extensions/r-wasm/live/_gradethis.qmd >}}

```{webr}
#| autorun: true
#| include: false
library(tidyverse)
```

## Introduction

So far, all data wrangling operations in this book operated 
on the rows, columns, or cells of data frames:
filtering the rows, selecting the columns, mutating or summarizing the values. 

This chapter expands on this by operating on the overall **shape** of data frames.
In the first section we will use `pivot` to turn columns into rows and the other way around.
Next, we will use `join` to combine multiple data sets together. 
We will use these techniques to look at the relation between global temperature rise and greenhouse gas emissions.  

<!-- 
Q: What we don't teach anywhere yet:
- bind_rows
- unnest (but will see that with strings in any case)
- separate / unite 
- replace_na
- rowwise
Do we need to teach this anywhere?
-->

### Data used in this chapter

This chapter will use data from the website [Our world in data](https://ourworldindata.org).
This is a very rich resource for various country-level facts and statistics. 
In particular we will use three data sets:

- [Global Temperature Anomaly](https://ourworldindata.org/grapher/temperature-anomaly),
  derived from the British [Met Office Hadley Centre HadCRUT5 data](https://www.metoffice.gov.uk/hadobs/hadcrut5/), 
  contains the yearly temperature from 1850 compared to a pre-industrial level baseline
  ([Open Government License](http://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/))
- [Greenhouse Gas Emissions](https://github.com/owid/co2-data), based on the 
  [Global Carbon Budget](https://globalcarbonbudgetdata.org/) (Global Carbon Project) and [National contributions to climate change](https://zenodo.org/records/7636699/latest) datasets,
  contains the CO2 and other greenhouse gas emissions per year and per country from 1850 (CC-BY license)
- [World Bank Income Groups](https://ourworldindata.org/grapher/world-bank-income-groups) based on
  the [World Bank country classification](https://datahelpdesk.worldbank.org/knowledgebase/articles/906519-world-bank-country-and-lending-groups)
  gives an annual classification from 1987 of countries into lower to high income groups  (CC-BY licensed)

We will be using smaller subsets of these data files that are redistributed along with this book.
If you're curious, you can see the [data sets documentation](https://github.com/vanatteveldt/r4css/blob/main/data/README.md) which contains the R code used to download and preprocess the files. 

## Pivotting data: wide vs long {#sec-pivot}
 
Let's take a look at the temperature anomaly dataset:

```{webr}
#| exercise: 'temp_data'
#| warning: false
library(tidyverse)
download.file("https://raw.githubusercontent.com/vanatteveldt/r4css/refs/heads/main/data/temperature_anomaly.csv", "temperature_anomaly.csv")
temp <- read_csv("temperature_anomaly.csv")
head(temp)
```

As you can see, this contains data per year for each 'region' (hemisphere or world average).
One row represents a combination of a year and a region,
with values for the temperature anomaly and the upper and lower confidence interval. 

Let's see if we can recreate a version of the famous 'barcode plot' on temperature rise: 

```{webr}
#| exercise: ['temp_barcode', 'temp_pivot_wider', 'temp_pivot_wider2']
#| setup: true
temp <- read_csv("data/temperature_anomaly.csv")
temp_diff <- temp |> 
  select(year, region, temp_anomaly) |> 
  pivot_wider(names_from=region, values_from=temp_anomaly) |>
  rename(south=`Southern Hemisphere`, north=`Northern Hemisphere`) |>
  mutate(diff=north - south) 
```
```{webr}
#| exercise: temp_barcode
ggplot(temp, aes(x=year, fill=temp_anomaly, y=fct_reorder(region, temp_anomaly))) + 
  geom_tile() + 
  scale_fill_gradient2(low="darkblue", mid="white", high="darkred", 
                       midpoint=.54, guide=NULL) + 
  labs(title="Temperature anomaly per hemisphere",
       x="",
       y="")
```

(Note that to produce the colors we used `scale_fill_gradient2`, which controls how
data values are mapped to colors (see @sec-advancedggplot for more information).
The neutral midpoint is set to .54, the average value for 1961-2010 also used in the original.)

### Long to wide data: `pivot_wider`

As you can see in the graph, the data is split by 'region', which contains rows for both hemispheres as well as for a global average. 
It is also clear that the northern hemisphere is warming up more quickly than the southern. 
However, if we want to calculate that gap we run into a problem: 
the values for north and south are in different rows, 
while we would need them in different columns on the same row to do a calculation.

In the lingo of tidyverse: we have **long data** (variables are in rows)
that we want to turn into **wide data** (variables are in columns). 

To do so, we use the function `pivot_wider`:

```{webr}
#| exercise: temp_pivot_wider
temp_diff <- temp |> 
  select(year, region, temp_anomaly) |> 
  pivot_wider(names_from=region, values_from=temp_anomaly) |>
  rename(south=`Southern Hemisphere`, north=`Northern Hemisphere`) |>
  mutate(diff=north - south) 
head(temp_diff)
```

As you can see, there is now one row per year,
with three data columns: one each for the unique values of the region
(`names_from=region`). 
The values represent the temperature anomaly in this year and region
(`values_from=temp_anomaly`). 

We can now plot this difference:

```{webr}
#| exercise: temp_pivot_wider2
temp_diff |>
  ggplot(aes(x=year, y=diff)) + 
  geom_hline(yintercept=0, color="red", alpha=.2) + 
  geom_line()
```

::: {.challenge title="Why the `select` function?"}

In the code example for `pivot_wider` above, we used a `select` to drop the confidence interval 
variables before pivotting. Why do you think we do this?

Remove the `select` line and re-run the example. 
Does the result make any sense? Why does it happen?
:::

As you saw when you removed the `select` line, this is a crucial step before pivoting.
The reason for this is that each column that is not used in the pivot (i.e. the `names_from` and `values_from` columns) 
are treated as identifying unique rows. So, each year is given a new row (as expected),
but without the select each `temp_lower` and `temp_upper` value is also seen as identifying a row!
So, the result is not one row per year, but one row per year, lower, and upper combination. 
Since the lower and upper bound tend to be unique, this means we create 3 rows per year.

The lesson here is: **before** doing `pivot_wider`, **always** make sure the data only contiains the columns
uniquely identifying the rows (here: `year`) and the names and values columns (`region` and `temp_anomaly`). 

```{r, eval=F}

### Using pivot_longer

# Just to teach the syntax, let's reverse the pivot_wider
pivot_longer(temp_diff, north:diff)
pivot_longer(temp_diff, -year, names_to="region", values_to="temperature_anomaly")


### Carbon dioxide data

# For a more serious use of pivot_longer let's have a look at the carbon data
co2  <- read_csv("data/co2_emissions.csv")
# Note: We remove 'world' from the csv so we can reinforce the groupby_summarize
co2 |> 
  group_by(year) |>
  summarize(co2=sum(co2, na.rm=TRUE)) |>
  ggplot(aes(x=year, y=co2)) + geom_line()

#  plot the co2, methane, no and ghg 
# This requires pivot_longer (to use color=variable)
ghg_sources <- co2 |>
  select(country, year, co2, methane, nitrous_oxide) |>
  pivot_longer(co2:nitrous_oxide, names_to="source")

ghg_sources |> 
  group_by(year, source) |>
  summarize(value=sum(value, na.rm=TRUE)) |>
  ggplot(aes(x=year, y=value, color=source)) + geom_line()

# Callout on 'tidy' data, esp for ggplot

### Exercise: compare two years
# e.g. (from claude:)

# NOTE: NEEDS SELECT
co2 |> filter(year %in% c(1990, 2020)) |>
  pivot_wider(names_from = year, values_from = co2, names_prefix = "y") |>
  mutate(change = 100 * (y2020 - y1990) / y1990)
# Might even be a good moment to demo a cleveland / difference lollipot chart

### Demo: dumbbell chart (from claude)

# geom_segment needs the start and end of each line in the same row,
# so this plot can only be made from the wide format
big_emitters <- co2 |>
  filter(year == 2020, country != "World") |>
  slice_max(co2, n = 16) |>
  pull(country)

per_capita <- co2 |>
  filter(year %in% c(1990, 2020), country %in% big_emitters) |>
  select(country, year, co2_per_capita) |>
  pivot_wider(names_from = year, values_from = co2_per_capita, names_prefix = "y") |>
  mutate(change = y2020 - y1990)

ggplot(per_capita, aes(y = fct_reorder(country, y2020))) +
  geom_segment(aes(x = y1990, xend = y2020, color = change > 0), linewidth = 1.2) +
  geom_point(aes(x = y1990), color = "grey50", size = 2) +
  geom_point(aes(x = y2020, color = change > 0), size = 3) +
  # Default colors map 'increase' to blue-green, which reads backwards for emissions
  scale_color_manual(values = c(`TRUE` = "firebrick", `FALSE` = "steelblue"), guide = "none") +
  labs(x = "CO2 emissions per capita (tonnes)", y = NULL,
       title = "Emissions per capita, 1990 (grey) to 2020")

# Challenge: the US falls from 20.2 to 13.9, India rises from 0.7 to 1.9.
# Which of the two is doing better? Does per capita or total emissions matter?
# (scale_color_manual is @sec-advancedggplot material - forward reference or drop the color)
```



## Joining multiple data sets

Claude: 

- 1:many blowup needs to be singalled explicitly
- coverage loss for income classification is a good explicit lesson
- order bar plot

```{r, eval=F}
# Are the rich or the poor countrires polluting?
# Let's get income class data
income <- read_csv("data/income_groups.csv")
income_2020 <- income |> filter(Year==2020) |> select(country=Entity, income=`World Bank's income classification`)

# Cutoff at 2020 and join
# Naive: inner join

co2_2020 <- co2 |> filter(year == 2020)
combined <- inner_join(co2_2020, income_2020) 
ggplot(combined, aes(y=income, x=cumulative_co2)) + geom_col()
# Note: Should summarize first!

# Callout (or challenge) on key columns and renaming/selecting (preferred) vs using by


### Explaining inner / full / left / right
combined_left <- left_join(co2_2020, income_2020) 
filter(combined_left, is.na(income))

### Explain semi and anti join as useful shorthand and debugging tools
anti_join(co2_2020, income_2020)
semi_join(co2_2020, income_2020)

# challenge to interpret this, especially venezuela,
# and figure out a way of dealing with it

### Exercise: co2 by country income type over time
income |> select(country=Entity, year=Year, income=`World Bank's income classification`) |>
  inner_join(co2) |>
  group_by(year, income) |>
  summarize(co2=sum(co2, na.rm=T)) |>
  ggplot(aes(x=year, y=co2, fill=income)) + geom_area()

# Maybe turn this into a demo+challenge instead?
# Challenge: why are low-income countries emitting less co2 over time?
# what is the bump in upper-middle countries in 2010? (answer: countries (esp china) are reclassified)
# How could you fix this? (answer: use a constant measure, e.g. use the 2020 classification instead)
# what changes if you use cumulative co2 instead? (answer: it makes the rich world's share clearer)
```

## Important take-away (callout)

for both joinign and pivoting the columns play an important role
this makes it very important to *first* select/rename to get the right columns
and then call the join or pivot
plus specific rules on what the 'right' columns are 

## Exercise: Are co2 emissions and temperature correlated over time?

(chapter-level exercise combining joining and pivoting)

```{r eval=F}
# Part A: Join
world_temp <- temp |> filter(region == "World")
world_co2 <- co2 |> group_by(year) |> summarize(co2=sum(cumulative_co2, na.rm=TRUE))
temp_and_co2 <- inner_join(world_temp, world_co2) 
cor.test(temp_and_co2$temp_anomaly, temp_and_co2$co2)

# Challenge: is the correlation actually meaningful? How to fix this?

# Part B: Pivot to plot
temp_and_co2 |> select(year, co2, temp_anomaly) |>
  pivot_longer(-year) |>
  group_by(name) |>
  mutate(index=value/mean(value) * 100) |>
  ggplot(aes(x=year, y=index, color=name)) + geom_line() 

```


